### 2025-09-01 docs/thesis_manuscript/manuscript.tex
generate plots now generates better resolution plots

Understanding this facet of evolution is [difficult]. Unlike with [allelic mutation], genome evolution occurs up to several orders of magnitude less freqently, and is more heterogenous in its manifestation. We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the [functional/coding] genome, stemming from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications, as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms are pseudogenization and deletions via various replication errors. here are some more specifics about the expansion/reduction process. here is why they are difficult to observe or create experiments to test. here is why is would be beneficial to create models that allow us to ask [] questions that bridge the gap between theory and experiment. 
Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications. (by doing so the cell acheives two things: 1. relieve pleiotropic effects, 2. lessen time to navigate sequence space). Alternatively, unused regions of the geome may also see genes spontaneously arise. (this process is much slower and harder to observe). In either case, these new genes are expected to accumulate neutrally, and over time have the opportunity to gain beneficial functionality. 
Although organisms are effective at removing unnecessary genes through pseudogenization, (this is a slow process, and still does not account for historical contingencies)
Different organisms may happen to adapt to the same environment in different way

We do however see in even shorter evolution experiments that not only mutation of alleles, but also genomic abberations appear that explain increases in fitness. (noted rate of nucleotide gain per generation stats here)
\subsection*{Fitness Landscape Theory}

As has been noted by studies of the \textit{de novo} evolution of genes, their emergence out of unused genetic substrate remains difficult to observe experimentally. (why? duplication vs de novo navigation of landscapes). There is thus a usefulness to generation of models that allow us to form hypotheses about these difficult to test systems.

Although fitness landscape theory has been used extensively to model adaptation on static genomes, it is less well developed on the scale of long term evolution, where changes in the genome size also arise. With this work we extend the Rough Mount Fuji model to incorporate genome evolution alongside allelic mutation

@@

@@

>>>>

Understanding this facet of evolution is difficult. Unlike with the point mutations that are typically observed in every generation, genome evolution occurs up to several orders of magnitude less frequently, and is more heterogenous in its manifestation. We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the functional genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications (WGD), as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors. Evolution has been observed to utilize both genome reduction and expansion in order to spur adaptation to increase the organism's \textit{fitness}, conceptualized as some metric representing success e.g. growth rate within an environment. However, more impactful processes such as neofunctionalization of duplications, de novo gene birth, and WGD, remain slow to induce and test, often in experiments running for years or decades. There is thus a usefulness to the creation of mathematical models that bridge the gap between theory and experiment and allow us to form hypotheses about these phenomena and their impact over long evolutionary timescales.\\

One such [question] naturally arising from the study of this type of adaptive system is regarding the effects of \textit{path dependency} during genome evolution: how does the order and timing in which evolutionary innovations arise affect evolutionary possibilities far in the future, potentially in different environments? For instance, the nitrogenase metalloenzyme - the main driver of bioavailable nitrogen cycling, which utilizes multiple oxygen-labile iron-sulfur clusters - emerged under the earth's early atmospheric conditions and favored adaptations suited to a reducing atmosphere. Faced with a later change to an oxidizing and highly destructive environment, organisms with this metabolic innovation, incapable of decommitting from fixing their own nitrogen to sustain cell growth, instead evolved a set of highly complicated cellular pathways to continue catalysis under only anoxic conditions, and protect the enzyme's metallic core from molecular oxygen. No other method of creating bioavailable nitrogen in an oxygenic cellular environment has ever been discovered. What if nitrogen fixation had first evolved after the oxygenation of the atmosphere? Is it even possible, and if so, how would the enzyme differ from today's? Evolution appears to be rife with examples of historical contingencies, spanning from within microbial metabolism up to macroscale physiology[cite giraffe]. [these are questions about the evolution of genomic context over time, and therefore a good starting point is from mathematical models that allow for the modeling of context, and how it can evolve. We thus turn to fitness landscapes.] how do different cellular contexts create different evolutionary outcomes, which then lead to even more different contexts, which lead to even different outcomes?

\subsection*{Fitness Landscape Theory}

The fitness landscape was first introduced in 19xx by Sewall Wright to illustrate xyz. With each unique combination of decisions there can be assigned a fitness value, and so we can generate a topology from a network of possible decision combinations. Fitnesses for contexts on the landscape may be higher or lower than expected from the sum of their individual components, leading to "ruggedness", otherwise known as \textit{epistasis}: nonadditivity resulting from higher-order interactions between components, that ultimately leads to, in the case of evolution, adapting populations reaching differing and suboptimal equilibrium points as they perform a "hill-climbing" style of optimization. This analogy has spurred a great deal of research in evolutionary biology, where it has found use in modeling adaptation of alleles, or more recently, in empirical fitness landscapes of proteins. The model can in fact be generalized to represent any sort of decision landscape, and has consequently found itself applied broadly in variety of fields such as [Directed evolution, economics?, and the other one]. Although fitness landscape theory has been used extensively to model adaptation on static genomes or within changing environments, it is less well developed on the scale of long term evolution, where changes in genomic context also arise and can be selected for. With this work we extend the Rough Mount Fuji model to incorporate genome evolution alongside allelic mutation to explore the effects of multiple mutational processes on evolutionary outcomes.

@@

We examine first the simple case where $\sigma_{e}$ is zero, resulting in a monopeaked landscape with no epistasis. Simulations were initialized with the following parameters, and xxx replicates were run by varying the seed value for the mutational rng.
\begin{center}
    \begin{tabular}{ | c | c | }
	\hline
	Maximum Size & 20 \\ \hline
	Initial Size & 1, 10, 20 \\ \hline
	Population & 1000 \\ \hline
	Generations & 50000 \\ \hline
	$\sigma_e$ & 0 \\ \hline
	$\sigma_a$ & 0.1 \\ \hline
	$\mu$  & 1e-4 \\ \hline
	$M$ & 1e-4 \\ \hline
    \end{tabular}
\end{center}

As there is no epistasis, populations will rapidly reach equilibrium at the fitness maximum on any individual landscape, whereas to find the global peak between all lanscapes a slighly different behavior is observed. Since a mutation to $0$ in either the genotype or genome produces an identical fitness, on average $50\%$ of negative additive effects will be removed via each type of mutation, and this balance will gradually fluctuate. Figure \ref{additivedynamics} shows genome sizes over time for starting genome sizes 1, 10, and 20 on a landscape with 10 positive additive effects.

\begin{figure}
	\centering
	\subfloat[1 locus initially accessible]{\includegraphics[width = 0.4\textwidth]{figures/L20l1i1_20250901_1230_cropped.pdf}}
	\label{fig:sub1}
	\subfloat[10 loci initially accessible]{\includegraphics[width = 0.4\textwidth]{figures/L20l10i1_20250901_1229_cropped.pdf}}
	\label{fig:sub2}
	\subfloat[20 loci initially accessible]{\includegraphics[width = 0.4\textwidth]{figures/L20l20i1_20250901_1229_cropped.pdf}}
	\label{fig:sub3}
	\subfloat[Densities of genome size time series]{\includegraphics[width = 0.4\textwidth]{figures/random_walk_additive.pdf}}
	\label{additivedynamics:sub4}
	\caption{caption}
	\label{additivedynamics}
\end{figure}

It can be seen that fitness reaches the same value in all three cases within relatively few generations, while genome size simultaneously moves to and thereafter fluctuates about 75\% of the maximum. This can be attributed to the random swapping of the two fitness-equivalent mutations as described above, while maintaining a $1$ state for all positive effects. In this example, the 10 positive additive effects will always be on, and on average, out of the remaining ten negative additive effects, five will be not present in the genome, while five will be and mutated to a $0$ state, resulting in a mean genome size of 15 averaged over the entire time series (Figure \ref{additivedynamics}d).

@@

As has been noted by studies of the \textit{de novo} evolution of genes, their emergence out of unused genetic substrate remains difficult to observe experimentally. (why? duplication vs de novo navigation of landscapes).

Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications. (by doing so the cell acheives two things: 1. relieve pleiotropic effects, 2. lessen time to navigate sequence space). Alternatively, unused regions of the geome may also see genes spontaneously arise. (this process is much slower and harder to observe). In either case, these new genes are expected to accumulate neutrally, and over time have the opportunity to gain beneficial functionality.
Although organisms are effective at removing unnecessary genes through pseudogenization, (this is a slow process, and still does not account for historical contingencies)
Different organisms may happen to adapt to the same environment in different way

Implications on the expanding genome
